Measuring causal influence with\\ back-to-back regression: the linear case

### 2019-11-12 paper/appendix.tex

\begin{figure}[t!]
  \centering
  \includegraphics[width=\textwidth, trim=0cm 0cm 0cm 0cm, clip=True]{figures/meg_supp.pdf}
  \caption{Average $\Delta R$ for each $Y$ dimension (Forward model), canonical components (PLS and CCA) or feature (B2B) across subjects. Formally, Backward model cannot have a $\Delta R$ because it never unmixes the multiple $X$ features. In such cases, we thus simply report the decoding score.}
  \label{fig:meg_supp}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{figures/AUC_R.png}
  \caption{Synthetic experiments. Distribution (over conditions) of differences in AUC (left) and Feature Importance $\Delta R$ (right) metrics between our method and the baselines. Mass under the dashed line corresponds to B2B outperforming the baseline. \label{fig:auc_plots}}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=0.7\linewidth]{figures/duration}
  \caption{Wall-clock run-time for our method B2B and for the baselines. We see that B2B runs much faster than baselines. \label{fig:duration}}
\end{figure}

>>>>

% \begin{figure}[t!]
%   \centering
%   \includegraphics[width=\textwidth, trim=0cm 0cm 0cm 0cm, clip=True]{figures/meg_supp.pdf}
%   \caption{Average $\Delta R$ for each $Y$ dimension (Forward model), canonical components (PLS and CCA) or feature (B2B) across subjects. Formally, Backward model cannot have a $\Delta R$ because it never unmixes the multiple $X$ features. In such cases, we thus simply report the decoding score.}
%   \label{fig:meg_supp}
% \end{figure}

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{figures/delta_auc.pdf}
  \caption{Synthetic experiments. Distribution (over conditions) of AUC (top) and Feature Importance $\Delta R$ (bottom) metrics between our method (y-axis) and the baselines (x-axis). Each dot is a distinct synthetic experiment. Dots below the diagonal indicates that B2B outperform the tested model. \label{fig:auc_plots}}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=0.7\linewidth]{figures/duration}
  \caption{Wall-clock run-time for our method B2B and for the baselines. Each dot is a distinct synthetic experiment. B2B runs much faster than cross-decomposition baselines. \label{fig:duration}}
\end{figure}

\subsection{Robustness on increasing number of factors}

To test whether each of the methods robustly scale to an increasingly large number of potential causes $X$, we enhance the four ad-hoc features (word length, word frequency, word function, dummy variable) with another ten features. These additional features corresponds to the first dimensions of word embedding as provided by Spacy \citep{spacy2}. The results shown in Figure \ref{fig:embbeding}, show that the ad-hoc features remain unchanged (and are actually improved) in the B2B.

\begin{figure}
  \centering
  \includegraphics[width=0.7\linewidth]{figures/compare_embeddings.pdf}
  \caption{Comparison of $\Delta R$ when the models are tested on four variables (top) and when the models are tested on an these four variables as well as another 10 word-embedding features (bottom). These results illustrate that, unlike Regularized CCA, B2B remains robust even when the number of tested factors increases.
  \label{fig:embeddings}}
\end{figure}

### 2019-11-12 paper/paper.tex

@@

% $$
\begin{equation}
  X^\top X = \left(\begin{array}{lccl}\Sigma_{X_1 X_1} & \Sigma_{X_1 X_2} \\ \Sigma_{X_1 X_2} & \Sigma_{X_2 X_2}\end{array}\right),\qquad\qquad N^\top N = \left(\begin{array}{lccl}\Sigma_{N_1 N_1} & \Sigma_{N_1 N_2} \\ \Sigma_{N_1 N_2} & \Sigma_{N_2 N_2}\end{array}\right),
\end{equation}
% $$
% and
\begin{equation}
  \hat H = \left(\begin{array}{cc}(\Sigma_{X_1 X_1}+\Sigma_{N_1 N_1})^{-1}\Sigma_{X_1 X_1} & (\Sigma_{X_1 X_1}+\Sigma_{N_1 N_1})^{-1}\Sigma_{X_1 X_2} \\0 & 0\end{array}\right),
\end{equation}
\begin{equation}
  Diag_k (\hat H) = Diag((\Sigma_{X_1 X_1}+\Sigma_{N_1 N_1})^{-1}\Sigma_{X_1 X_1}) = Diag((I+\Sigma_{X_1 X_1}^{-1}\Sigma_{N_1 N_1})^{-1}).
\end{equation}
%
% $$
% $$
% $$
% $$
%
In the absence of noise, we have $\Sigma_{N_1 N_1}=0$, and so $Diag_k(\hat H)=I$, and $$Diag(\hat H) = Diag(E)$$

Otherwise the $k$ first diagonal elements of $\hat H$ are all positive (since we assume that $F$ has full rank over $Img(E)$, they cannot be zero), lower than $1$, and bounded by 
$$\frac{\sigma_{X_k}}{\sigma_{X_k} +\sigma_{N_1}} \text{ and }  \frac{\sigma_{X_1}}{\sigma_{X_1} +\sigma_{N_k}}$$where  $\sigma_{X_1}$, $\sigma_{X_k}$, $\sigma_{N_1}$ and $\sigma_{N_k}$ denote the largest and smallest eigenvalues of $\Sigma_{X_1 X_1}$ and $\Sigma_{N_1 N_1}$.
In real cases, noise will be added to the diagonal elements of $\hat H$, by non-linearities and sampling error.

The average value of the coefficients of $Diag_k(\hat H)$ is the trace of $\hat H$ divided by $k$. Since $X$ and $N$ are not correlated, $\Sigma_{X_1 X_1}^{-1}\Sigma_{N_1 N_1}$ should have a low condition number, and $\frac{Var(X)}{Var(X)+Var(N)}$ provides a good estimate of $\frac{1}{k}Tr(\hat H)$, that could be used to test whether a feature of $X$ has causal influence. The Sonquist and Morgan criterion defined in the appendix is another possibility.

>>>>

Note that the absence, in our model, of measurement noise over $Y$, is not restrictive, so long $F$ has full-rank (or over $Img(F)$, the subspace spanned by the columns of $F$). In fact, adding $M$ as measurement noise on $Y$, we have $$Y=(XE+N)F+M$$ On $Img(F)$, 
rewriting $M = MF^{-1}F$, we have
$$Y = (XE+N)F+M = (XE+N+MF^{-1})F = (XE+N')F$$

@@

% $$
\begin{equation}
  X^\top X = \left(\begin{array}{lccl}\Sigma_{X_1 X_1} & \Sigma_{X_1 X_2} \\ \Sigma_{X_1 X_2} & \Sigma_{X_2 X_2}\end{array}\right),\qquad\qquad N^\top N = \left(\begin{array}{lccl}\Sigma_{N_1 N_1} & \Sigma_{N_1 N_2} \\ \Sigma_{N_1 N_2} & \Sigma_{N_2 N_2}\end{array}\right),
\end{equation}
% $$
% and
\begin{equation}
  \hat H = \left(\begin{array}{cc}(\Sigma_{X_1 X_1}+\Sigma_{N_1 N_1})^{-1}\Sigma_{X_1 X_1} & (\Sigma_{X_1 X_1}+\Sigma_{N_1 N_1})^{-1}\Sigma_{X_1 X_2} \\0 & 0\end{array}\right),
\end{equation}
\begin{equation}
  Diag_k (\hat H) = Diag((\Sigma_{X_1 X_1}+\Sigma_{N_1 N_1})^{-1}\Sigma_{X_1 X_1}) = Diag((I+\Sigma_{X_1 X_1}^{-1}\Sigma_{N_1 N_1})^{-1}).
\end{equation}
%
% $$
% $$
% $$
% $$
%
In the absence of noise, we have $\Sigma_{N_1 N_1}=0$, and so $Diag_k(\hat H)=I$, and $$Diag(\hat H) = Diag(E)$$ 
Otherwise the $k$ first diagonal elements of $\hat H$ are all positive (since we assume that $F$ has full rank over $Img(E)$, they cannot be zero), lower than $1$, and bounded by 
$$\frac{\sigma_{X_k}}{\sigma_{X_k} +\sigma_{N_1}} \text{ and }  \frac{\sigma_{X_1}}{\sigma_{X_1} +\sigma_{N_k}}$$where  $\sigma_{X_1}$, $\sigma_{X_k}$, $\sigma_{N_1}$ and $\sigma_{N_k}$ denote the largest and smallest eigenvalues of $\Sigma_{X_1 X_1}$ and $\Sigma_{N_1 N_1}$.

In practical cases, non linearities and sampling error will add noise to the diagonal elements of $\hat H$. To test diagonal elements for equality with zero, we note that the average value of the non-zero coefficients of $Diag(\hat H)$ is the trace of $\hat H$ divided by $k$. Since $X$ and $N$ are not correlated, $\Sigma_{X_1 X_1}^{-1}\Sigma_{N_1 N_1}$ should have a low condition number, which means $\frac{Var(X)}{Var(X)+Var(N)}$ should provide a good estimate of $\frac{1}{k}Tr(\hat H)$. This can be used to test whether a feature of $X$ has causal influence. An alternative is the Sonquist-Morgan criterion presented in the appendix.
